---
title: "eDNA explorer"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F, warning=F)

# load libraries ----
suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
  library(lubridate)
  library(ggplot2)
  library(plotly)
  library(RColorBrewer)
  library(crosstalk)
  library(leaflet)
  library(DT)
})

# load data ----
# setwd('docs') # for debug, not knitting

if (interactive() & basename(getwd()) != 'docs'){
  setwd('docs')
}

otu_csv   = '../data/otu.csv'
otl_csv   = '../data/otl.csv'
sites_csv = '../data/sites.csv'

otu   = read_csv(otu_csv)
otl   = read_csv(otl_csv)
sites = read_csv(sites_csv)

# Wrap data frame in SharedData
otu_ms = otu %>%
  mutate(
    month = month(date),
    date  = as_date(date)) %>%
  left_join(sites, by=c('site'='site_code'))
```

```{r crosstalk filter by date}
# shared data
sd = otu_ms  %>%
  group_by(site, site_name, lon, lat, date, month) %>%
  summarize(
    phylum = 'ALL',
    n_otu = sum(count)) %>%
  ungroup() %>%
  SharedData$new()

# filters
filter_slider(
  'date', 'Date Sampled', sd, column=~date)

# color palette
pal <- colorNumeric('YlOrRd', sd$data()$n_otu) # 'Spectral'

# arrange in bootstrap columns
bscols(
  
  # map of sites
  leaflet(sd) %>% 
    addProviderTiles('Esri.OceanBasemap') %>%
    addCircleMarkers(
      radius = ~rescale(n_otu, to=c(0.1, 10)),
      color = ~pal(n_otu),
      stroke = FALSE, fillOpacity = 0.5) %>%
    addLegend("topleft", pal = pal, values = sd$data()$n_otu,
      title = "OTUs", opacity = 1),
  
  # plot of otus over time
  plot_ly(data = sd, x = ~date, y = ~n_otu, color = ~site) %>%
    add_lines() %>%
    layout(xaxis = list(title='Month'), yaxis = list(title='OTUs')))

# table
datatable(
  sd, extensions="Scroller", style="bootstrap", class="compact", width="100%",
  colnames=c(
    'Site'='site','Name'='site_name','Longitude'='lon','Latitude'='lat',
    'Date'='date','Month'='month','OTUs'='n_otu'),
  options=list(deferRender=TRUE, scrollY=300, scroller=TRUE))
```

## Notes

### Preparation

See [prep.Rmd]().

### Interactivity

You can interactively filter and view data above by:

- **date slider**. Drag slider to define date range to display.
- **map box**: Click on box button in upper left to limit range.
- **plot legend**: Click on legend labels to turn on/off sites in plot and hover mouse to get values for date.
- **table row**: Click on rows to highlight selection in map and plot.

### Issues

Issues:

- **map**. Markers at each site are plotted for each date sampled, whereas they should be aggregated across the filters selected (requires Shiny).
- **taxa**. Taxonomic filters are not feasible without aggregating also by site and/or date (requires Shiny).


### Next Steps

Experimenting with time-space-taxa filtering and display using [Crosstalk](http://rstudio.github.io/crosstalk/) which enables knitting from [Rmarkdown](http://rmarkdown.rstudio.com/) (see [source]()) to HTML for free easy hosting on [Github Pages](https://pages.github.com/).

Although Crosstalk synchronizes a shared data object between [htmlwidgets](http://www.htmlwidgets.org/) such as the [leafet]() map, [plotly]() plot, and [DT]() table, the need for aggregation across time, taxa and/or sites, necessitates migrating this type of application to [Shiny](http://shiny.rstudio.com/) for applying back-end R functions to aggregate data as interactively specified by the user.

Shiny requires a hosted solution with [Shiny Server](https://www.rstudio.com/products/shiny/shiny-server/) or [ShinyApps.io](http://www.shinyapps.io/). I'll proceed with installing the open-source Shiny Server in the [Google Cloud Platform](https://cloud.google.com/).