---
title: "eDNA Explorer with crosstalk"
#date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: false
    toc_depth: 3
---

The [crosstalk](http://rstudio.github.io/crosstalk/) package allows filtering and synchronization between htmlwidgets (ie map, plot and table) on a shared data object. Since it does not require an R backend server to render functions like Shiny, it can be hosted for free on any website like Github Pages.

There are however severe limitations in this case, since we want to aggregate across sites or times.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F, warning=F)

# load libraries ----
suppressPackageStartupMessages({
  library(tidyverse)
  library(htmltools)
  library(scales)
  library(lubridate)
  library(ggplot2)
  library(plotly)
  library(RColorBrewer)
  library(crosstalk)
  library(leaflet)
  library(DT)
})

# load data ----
# setwd('docs') # for debug, not knitting

if (interactive() & basename(getwd()) != 'docs'){
  setwd('docs')
}

otu_csv   = '../data/otu.csv'
otl_csv   = '../data/otl.csv'
sites_csv = '../data/sites.csv'

otu   = read_csv(otu_csv)
otl   = read_csv(otl_csv)
sites = read_csv(sites_csv)

# Wrap data frame in SharedData
otu_ms = otu %>%
  mutate(
    month = month(date),
    date  = as_date(date)) %>%
  left_join(sites, by=c('site'='site_code'))
```

```{r crosstalk filter by date}
# shared data
sd = otu_ms  %>%
  group_by(site, site_name, lon, lat, date, month) %>%
  summarize(
    phylum = 'ALL',
    n_otu = sum(count)) %>%
  ungroup() %>%
  SharedData$new()

# filters
filter_slider(
  'date', 'Date Sampled', sd, column=~date)

# color palette
pal <- colorNumeric('YlOrRd', sd$data()$n_otu) # 'Spectral'

# arrange in bootstrap columns
bscols(
  
  # map of sites
  leaflet(sd) %>% 
    addProviderTiles('Esri.OceanBasemap') %>%
    addCircleMarkers(
      radius = ~rescale(n_otu, to=c(0.1, 10)),
      color = ~pal(n_otu),
      stroke = FALSE, fillOpacity = 0.5) %>%
    addLegend("topleft", pal = pal, values = sd$data()$n_otu,
      title = "OTUs", opacity = 1),
  
  # plot of otus over time
  plot_ly(data = sd, x = ~date, y = ~n_otu, color = ~site) %>%
    add_lines() %>%
    layout(xaxis = list(title=''), yaxis = list(title='OTUs')))

p()

# table
datatable(
  sd, extensions="Scroller", style="bootstrap", class="compact", width="100%",
  colnames=c(
    'Site'='site','Name'='site_name','Longitude'='lon','Latitude'='lat',
    'Date'='date','Month'='month','OTUs'='n_otu'),
  options=list(deferRender=TRUE, scrollY=300, scroller=TRUE))
```

## Notes

### Preparation

See [prep](./prep.html) (source [prep.Rmd](https://github.com/marinebon/edna-vis/blob/master/docs/prep.Rmd)) on manipulations to go from original [OTU_table_taxa_all.txt](https://github.com/marinebon/edna-vis/blob/master/data/OTU_table_taxa_all.txt) to the [otu.csv](https://github.com/marinebon/edna-vis/blob/master/data/otu.csv) used in this visualization.

You can see the Rmarkdown source for this doc at [index.Rmd](https://github.com/marinebon/edna-vis/blob/master/docs/index.Rmd).

### Interactivity

You can interactively filter and view data above with:

- **date slider**. Drag slider to define date range to display.
- **map box**: Click on box button in upper left to limit range.
- **plot legend**: Click on legend labels to turn on/off sites in plot and hover mouse to get values for date.
- **table row**: Click on rows to highlight selection in map and plot.

### Issues

Issues:

- **map**. Markers at each site are plotted for each date sampled, whereas they should be aggregated across the filters selected (requires Shiny).
- **taxa**. Taxonomic filters are not feasible without aggregating also by site and/or date (requires Shiny).

### Next Steps

Experimenting with time-space-taxa filtering and display using [Crosstalk](http://rstudio.github.io/crosstalk/) which enables knitting from [Rmarkdown](http://rmarkdown.rstudio.com/) (see [source]()) to HTML for free easy hosting on [Github Pages](https://pages.github.com/).

Although Crosstalk synchronizes a shared data object between [htmlwidgets](http://www.htmlwidgets.org/) such as the [leafet](http://rstudio.github.io/leaflet/) map, [plotly](https://plot.ly/r/) plot, and [DT](http://rstudio.github.io/DT/) table, the need for aggregation across time, taxa and/or sites, necessitates migrating this type of application to [Shiny](http://shiny.rstudio.com/) for applying back-end R functions to aggregate data as interactively specified by the user.

Shiny requires a hosted solution with [Shiny Server](https://www.rstudio.com/products/shiny/shiny-server/) or [ShinyApps.io](http://www.shinyapps.io/). I'll proceed with installing the open-source Shiny Server in the [Google Cloud Platform](https://cloud.google.com/).