---
title: "Prep eDNA Data"
author: "Ben Best"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
categories: ["R"]
tags: ["R Markdown", "eDNA", "prep"]
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(lubridate)
library(DT)
library(leaflet)
```

# Parse OTU

Parse:

- from [OTU_table_taxa_all.txt](https://github.com/marinebon/edna-vis/blob/master/data/OTU_table_taxa_all.txt): original tab-seperated text file in wide format (OTU rows x site columns)

- to [otu.csv](https://github.com/marinebon/edna-vis/blob/master/data/otu.csv): comma-seperated file with zeros and other data filtered out (not having site or taxanomic identification) in long format (OTU, site rows)

```{r read otu, warning=F, message=F}
otu_txt = '../data/OTU_table_taxa_all.txt'
otu_csv = '../data/otu.csv'

# setup columns for otu
otu_1 = read_tsv(otu_txt, n_max=1)
otu_cols = c(names(otu_1), 'taxa_v')

# read otu, in wide format with many 0s, and extra taxa vector column named
otu_w = read_tsv(otu_txt, col_names=otu_cols, skip=1)

# gather into long format, sparse 0s removed, column names and values expanded
otu = otu_w %>%                                                                   #   5,707 rows
  gather(sample, count, -DUP_ID, -taxa_v) %>%                                     # 291,057 rows
  filter(count > 0) %>%                                                           #  18,616 rows
  filter(!is.na(taxa_v)) %>%                                                      #   2,097 rows
  filter(
    !sample %in% c('extr_blank_1','extr_blank_2','extr_blank_3','Negative')) %>%  #   2,082 rows
  rownames_to_column('row_long') %>%
  extract(
    sample, 
    c('site','moyr','replicate_ltr','replicate_num'),
    '([A-Z]{2})([0-9]{3,4})([a-z]{1})([0-9]{0,3})',
    remove=T, convert=T) %>%
  mutate(
      date = parse_date(sprintf('01%04d', moyr), '%d%m%y')) %>%
  select(-moyr) %>%
  extract(
      taxa_v, 
      c('kingdom','phylum','class','order','family','genus','species'), 
      "'k__(.*)','p__(.*)','c__(.*)','o__(.*)','f__(.*)','g__(.*)','s__(.*)'",
      remove=T)

write_csv(otu, otu_csv)

datatable(otu)
```

# Data Questions

## What are the lon/lat for sites?

```{r}
table(otu$site, useNA='ifany')
```



```{r sites csv}
sites_csv = '../data/sites.csv' # sites_csv = 'data/sites.csv'

sites = tribble(
  ~site_code, ~site_name,      ~lon,        ~lat,
  'MR'      , 'Molasses Reef', -81.4317097, 24.5524431,
  'LK'      , 'Looe Key',      -80.3812147, 25.0165408,
  'WS'      , 'Western Sambo', -81.713333, 24.477778)

write_csv(sites, sites_csv)

datatable(sites)
```


```{r sites map}
leaflet(sites) %>%
  addProviderTiles('Esri.OceanBasemap') %>%
  addMarkers(
    ~lon, ~lat, 
    popup = ~sprintf("%s: %s", site_code, site_name))
```



## What's the meaning of a "t#" replicate vs others?

```{r}
table(otu$replicate_ltr, otu$replicate_num, useNA='ifany')
```

## Are we OK with getting rid of non-site data?

Eg columns: ``r c('extr_blank_1','extr_blank_2','extr_blank_3','Negative')``

## Are we OK to skip OTUs without any identifiable taxa?

Per code above applying `filter(!is.na(taxa_v))` goes from 18,616 to 2,097 rows.
